pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-07-02 bmc_article.tex
Updates from Overleaf

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}. We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.

We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). MCVR did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), this was at the expense of others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by MCVR and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

Of note, the optimal method in our evaluation (\textit{maxLikGlobal}) systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. {\color{red}Unsupervised clustering of the results reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results. This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of Seurat's resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 28A). Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B). We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}). The combination of \textit{scran} normalization with the elbow method for selection the number of dimensions appeared to offer the worst results. The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset instead standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}

We next investigated whether the benefits of some methods were conditional on the choices at other steps. Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 31A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 31B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{DCA} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:

@@

Our recommendations are in line with the results of several other benchmarking studies. For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}. A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVERX} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations. Another disagreement with the present recommendations come from the study of Hou et al. \cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools. Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies. \textit{ALRA} however yielded good results in both studies and may be the most conservative option. The study from Hou et al. also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies. The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}. One of the main strengths of \textit{scVI} is its scalability to bigger datasets as the model can be enhanced by a bigger training set. The model was thus reported as less adapted to datasets with a higher number of genes compared to their number of cells. As all of our benchmarking datasets fall in this category, it is very likely that the poor performance of \textit{scVI} in our evaluation is a consequence of the poorer training that \textit{scVI} can perform. For instance, the original article could train their model on a PBMC dataset of 12'039 cells whereas we used a FACS-sorted version of the same experiment, down-sampled to only 6'498 cells at maximum. Another limitation is that the imputation and normalization performed by \textit{scVI} were initially not designed to improve cluster detection. On the opposite to the low-dimensional space returned by the model was intended for this purpose and indeed showed a higher ranking compared to other methods, as demonstrated in the original article [NOTE TO MYSELF; adapt to results. For instance Hou2020 found that scVI was not among the top performers wrt to clustering based on latent space. And Aparicio 2020 that found a low silhouette score using scVI's imputations as an input for clustering].

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make 
clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

@@

\subsection*{\textit{scVI} normalization and LD} [NOTE TO MYSELF; adapt to the new runs]

>>>>

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI}-based normalization \cite{lopezDeep2018} was included separately{\color{red}, as it was not meant for this purpose and follows a slightly different flow}). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. {\color{red}An exception was \textit{scVI}-based normalization, which performed similarly to other methods on some datasets (Kumar, simMix2, Zhengmix4eq, Zhengmix4uneq), but very poorly on others, in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; of note, these values were not meant to be used for clustering, and the usage of the \textit{scVI} latent space will be discussed below along with dimensionality reductions)}. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

The impact of the dimensionality reduction methods tested was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. {\color{red} Like GLM-PCA, \textit{scVI}'s Linearly Decoded (LD) data \cite{SvenssonLDVAE2020} and latent space do not explicitly rely on normalized counts. Their performance was however worse than \textit{Seurat}'s PCA in view of the silhouette width (Supplementary Figure 15). In terms of clustering accuracy \textit{scVI}'s latent space performed better that its LD but was nevertheless less accurate than PCA-based alternatives on the datasets with a very low amount of cells (Koh and Kumar) and harboring many subpopulations (simMix1 and Zhengmix8eq).}

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}. {\color{red} We also included a variation of the molecular cross-validation (MCVR, modified to be applied on normalized data - see methods) approach originally developed to select an calibrate denoising methods \cite{batsonMolecularCrossValidationSingleCell2019}, but which has also been applied to estimate dimensionality \citep{wagnerMonetOpensource2020}.}

We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{sctransform} or standard \textit{Seurat} normalization.

{\color{red}\textit{MCVR} and the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}) tended to produce clusterings with smaller deviations to the true number of subpopulations (Supplementary Figure 26). Although most methods performed well on the various clustering measures, \textit{maxLikGlobal} (using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k} -- see Supplementary Figure 27) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). \texit{MCVR} did not perform well for the datasets without UMIs (Koh and Kumar), which violate the assumptions of the method, and while it did increase the separability of some subpopulations (Figure \ref{fig:dim}C), it was detrimental to the classification of some others. We observed a strong negative correlation (\textit{r}=-0.995, p~1.6e-8) between the number of dimensions called by \textit{MCVR} and the log10 average library size of a dataset, which explained the former considerably better than the number of subpopulations.

Of note, the optimal method in our evaluation (\textit{maxLikGlobal})} systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. {\color{red}Unsupervised clustering on the evaluation metrics reveals general patterns in the results as well as the relative importance of the parameters (Figure \ref{fig:overall}), with the clustering method having the most important impact on the results. This is however chiefly due to number of clusters called: an ANOVA on all analyses after accounting, in the model, for baseline differences between datasets and the impact of \textit{Seurat}'s resolution parameter, attributed relatively little importance to the clustering method, and instead attributed the largest proportion of variance in the main cluster metrics to the number of dimensions used, followed by normalization (Supplementary Figure 29A). Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 29B). We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality (as well as feature selection) came closely behind (Supplementary Figure 29B), suggesting that more attention should be paid to this relatively neglected step.

The greater importance of normalization, dimensionality and feature selection over filtering steps is also visible in the clustering of (Figure \ref{fig:overall}). The top methods (bottom of Figure \ref{fig:overall}) confirm the results of the previous sections, highlighting especially \textit{sctransform}, deviance-based feature selection, and \textit{maxLikGlobal} dimensionality estimates as top methods. Interestingly, while the combination of \textit{sctransform} with \textit{Seurat} clustering was particularly performant for most datasets, for the simulated `simMix1` dataset instead standard \textit{Seurat} normalization combined with \textit{scran}-based clustering proved superior.}

We next investigated whether the benefits of some methods were conditional on the choices at other steps. Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 30-31). For instance, we confirmed that doublet removal and \textit{sctransform} normalization were beneficial independently of other steps, and that, again independently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 30). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 31). {\color{red} ANOVA did reveal some significant (negative) interactions: the top two-way interaction was between \textit{scran} normalization and the elbow method for selection the number of dimensions, and indeed that combination was involved in all of the 30 lowest-ranking pipelines. Furthermore, all significant interactions involved the elbow method, suggesting that some methods were less robust to underestimated dimensionality.}

@@

The basic pipeline presented here can be extended with additional analysis steps while keeping the same evaluation metrics. To demonstrate this, we evaluated various imputation or denoising techniques based on their impact on classification. Since preliminary analysis showed that all methods performed equally well or better on normalized data, we applied them after filtering and normalization, but before scaling and reduction. Although some of the methods (e.g., \textit{DRImpute\_process} and \textit{alra\_norm}) did improve the separability of some more elusive subpopulations, no method had a systematically positive impact on the average silhouette width across all subpopulations (Supplementary Figure 32A). When restricting ourselves to clustering analyses that yielded the `right' number of clusters, all tested methods improved classification  compared to a scenario with no imputation step (`none' label, Supplementary Figure 32B). However, the situation was not so straightforward with alternative metrics, where some methods (e.g., \textit{ENHANCE}) consistently underperformed. 10X datasets, which are typically characterized by a lower per-cell coverage and feature detection rate, benefited more from imputation and, in this context, \textit{DrImpute} and \textit{DCA} tended to show the best performance. On the contrary, imputing on normalized counts originating from Smart-seq technology was instead rather deleterious to the clustering accuracy.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 33-34). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 35:

@@

Our recommendations are in line with the results of several other benchmarking studies. For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}. A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVERX} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations. Another disagreement with the present recommendations come from the study of Hou et al. \cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools. Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies. \textit{ALRA} however yielded good results in both studies and may be the most conservative option. The study from Hou et al. also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.

{\color{red} The performance of \textit{scVI} in our evaluation was lower than in the original publication presenting the method \cite{lopezDeep2018}. Our results are however in line with other studies comparing silhouette score \cite{Aparicio2020} and clustering accuracy \cite{HouImput2020} of \textit{scVI}'s latent space. However, these studies, like ours, used datasets with relatively few cells (i.e. fewer than genes), for which \textit{scVI} was reported by its authors to be less adapted. Further work will therefore be required to assess compare its performance on larger datasets.}

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make 
clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 35). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation. {\color{red}All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).}

@@

\subsection*{\textit{scVI} normalization and LD}

The evaluation of \textit{scVI} algorithm was performed separate to the other pipeline combinations as the model training on our reference datasets was especially resource-consuming {\color{red} and follows a different flow from the other pipelines. In particular, for datasets with fewer cells than genes, \textit{scVI} requires a feature selection to be performed prior to the normalization and imputation steps; furthermore, its dimensional reduction steps rely directly on raw counts. We therefore evaluated it separately and compared it to a representative subset of pipelines using \textit{sctransform} normalization and \textit{Seurat}'s PCA dimension reduction method. The resulting evaluation is available in the Supplementary Figure 13.
